---
title: "Liver 5"
author: "Martina Bonomi"
date: "2023-02-14"
output: 
  html_document:
    number_sections: yes
    self_contained: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(results='asis')
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_knit$set(root.dir = getwd())

library(ggplot2)
library(tidyverse)
library(Matrix)
library(scales)
library(cowplot)
library(RCurl)
library(ggplot2)
library(devtools)
library(reconPlots)
library(png)
library(knitr)
library(readxl)

text_size=5

```


```{r}

# read a csv file for a plan from fiona's output and create a list of two objects: one object is the dataframe with dvhs, the other one is a dataframe with volumes [cc] for each dvh


read_plan <- function(csv_file){
  
  # ---------------------------------------------------- #
  # csv_file -> csv file to read, FIonA output
  # ---------------------------------------------------- #
  
  # read csv file
  dvh <- read.csv(csv_file)
  colnames(dvh)[1] <- "Dose"
  # create df with volumes [cc] from last row of plan
  vol_cc <- tail(dvh, n=1)
  vol_cc <- vol_cc[2:length(dvh)]
  # keep just dvh data
  dvh <- dvh[1:nrow(dvh)-1,]
  # convert dose to num (from chr)
  dvh$Dose <- as.numeric(dvh$Dose)
  # get organs I have
  organs <- str_to_title(colnames(dvh)[4:length(dvh)])
  # create plan > list of dvh and vol 
  plan <- list(dvh,vol_cc, organs)
  names(plan) <- c("DVH data", "volumes [cc]", "organs")
  return(plan)
}

# ----------------------------------------------------------------------------- #

get_organs <- function(plan){
  organs <- plan[["organs"]]
  return(organs)
}

# ----------------------------------------------------------------------------- #

# find dose or volume values 

find_Dv <- function(plan, v, organ){
  
  # ---------------------------------------------------- #
  # plan -> plan for which you want to find the Dv
  # v [num] -> volume [%]
  # organ -> organ for which you want to compute Dv
  # ---------------------------------------------------- #
  
  # get names of DVHs in the plan file
  dvhs <- colnames(plan[["DVH data"]])
  
  # get column idx for wanted organ
  organ_idx <- which(str_detect(dvhs), toupper(organ))
  
  # get organ data and dose values
  organ_data <- data.frame(x = plan[["DVH data"]]$Dose , y = plan[["DVH data"]][organ_idx])
  
  # draw an horizontal line to get the intersection between v and the dvh
  vol <- data.frame(x = c(0, 10), y = c(v, v))
  
  # get intersection
  intersection <- curve_intersect(organ_data,vol)
  
  # return Dv value
  return(intersection[[1]])

}

# ----------------------------------------------------------------------------- #

find_Vd <- function(plan, d, organ){
  
  # ---------------------------------------------------- #
  # plan -> plan for which you want to find the Vd
  # d [num] -> dose [%]
  # organ -> organ for which you want to compute Vd
  # ---------------------------------------------------- #
  
  # get names of DVHs in the plan file
  dvhs <- colnames(plan[["DVH data"]])
  
  # get column idx for wanted organ
  organ_idx <- which(str_detect(dvhs), toupper(organ))
  
  # get idx for wanted dose
  dose_idx <- which(plan[["DVH data"]]$Dose == d)
  
  # get Vd
  value <- plan[["DVH data"]][organ_idx]
  
  return(value)
}

# ----------------------------------------------------------------------------- #

# Plot different DVHs for different plans or underdoses/overdoses

plot_DVHs <- function(plans, plan_names, dvhs_stat, title){
  
  # --------------------------------------------------------------------------------- #
  # plans [list] -> list of plans you want to compare by plotting them (even one)
  # plan_names [chr] <- vector of names of your plans
  # dvhs_stat [logic] -> vector of conditions for dvhs you want to plot (either TRUE or FALSE)
  # --------------------------------------------------------------------------------- #
  
  # get names of plans and DVHs
  names(plans) <- plan_names
  dvhs <- toupper(names(dvhs_stat))
  
  dvhs_list <- list()
  dvhs_df <- list()
  
  # loop over plans to get wanted DVHs to plot
  for(plan in plan_names){
    
    # list of dvhs you actually want to plot
    dvhs_list[[plan]] <- list()
    dvhs_list[[plan]][["Dose"]] <- plans[[plan]][["DVH data"]][,1]
    
    # loop over DVHs to select the ones which are set to TRUE
    for(dvh_idx in 1:length(dvhs)){
      
      # if the organ is TRUE, then I add it to the plot
      if(dvhs_stat[[dvh_idx]]==TRUE){
        
        # select position of the organ in the dvh dataframe
        idx <- which(str_detect(colnames(plans[[plan]][["DVH data"]]), dvhs[dvh_idx]))
        
        # selecting current dvh in the dataframe and current name
        if(dvhs[dvh_idx] == "PTV" || dvhs[dvh_idx] == "CTV" ){
          curr_dvh_name <- as.character((dvhs[dvh_idx]))
        } else {
          curr_dvh_name <- as.character(str_to_title(dvhs[dvh_idx]))
        }
        
        # selecting data of current DVH
        curr_dvh <- plans[[plan]][["DVH data"]][,idx]
        
        # adding current dvh to the dvhs list
        dvhs_list[[plan]][[curr_dvh_name]] <- curr_dvh
      }
    }
    
    dvhs_list[[plan]] <- as.data.frame(dvhs_list[[plan]])
    
    # gathering together DVHs to plot 
    dvhs_df[[plan]]  <- dvhs_list[[plan]]  %>% 
      gather(key = "DHV_curve", value = "value", -Dose)
    
    # add label to the plan
    dvhs_df[[plan]]$Plan <- plan
    
  }
  
  colors <- c("cornflowerblue","violetred", "orange", "chartreuse","deeppink3", "darkturquoise", "mediumorchid1")
  
  # create plot grid
  plot <- ggplot(dvhs_df[[plan_names[1]]], aes(x=Dose, y=value)) + 
    ylim(0,105) + 
    xlab("Dose [%]") +
    ylab("Volume [%]") 
  
  if(title == TRUE){
    # create plot title
    title <- c()
    for(i in 1:length(plan_names)){
      if(i==length(plan_names)){
        title <- paste(title, plan_names[i], sep=" ")
      } else{
        title <- paste(title, plan_names[i], "vs.", sep=" ")
      }
    }
    plot <- plot + ggtitle(title)
  } else if(title == FALSE){
    plot <- plot
  } else{
    plot <- plot + ggtitle(title)
  }
  
  # plot DVHs for each plan
  for(plan in plan_names){
    if(str_detect(plan, "underdose")==TRUE || str_detect(plan, "overdose")==TRUE){
      plot <- plot + geom_line(data=dvhs_df[[plan]], aes(color = DHV_curve, linetype=Plan), linewidth=0.4)
    }else{
      plot <- plot + geom_line(data=dvhs_df[[plan]], aes(color = DHV_curve, linetype=Plan), linewidth=0.8)
    }
  }
  
  plot <- plot + scale_color_manual(name = "DVH curves", values = colors)
  
  return(plot)
  
}

# ----------------------------------------------------------------------------- #

# plot dvhs for different configurations (one dvh at a time)

plot_DVH_comparison <- function(plans, plan_names, dvh, title){
  
  # --------------------------------------------------------------------------------- #
  # plans [list] -> list of plans you want to compare by plotting them (even one)
  # plan_names [chr] <- vector of names of your plans
  # dvh -> dvh you want to plot in comparison with the others
  # --------------------------------------------------------------------------------- #
  
  # get names of plans and dvh
  names(plans) <- plan_names
  dvh <- toupper(dvh)
  
  dvh_list <- list()
  dvh_df <- list()
  
  # loop over plans to get wanted dvh to plot
  for(plan in plan_names){
    
    # list of dvh you actually want to plot
    dvh_list[[plan]] <- list()
    dvh_list[[plan]][["Dose"]] <- plans[[plan]][["DVH data"]][,1]
    
      
    # select position of the organ in the dvh dataframe
    idx <- which(str_detect(colnames(plans[[plan]][["DVH data"]]), dvh))
      
    # selecting current dvh in the dataframe and current name
    if(dvh == "PTV" || dvh == "CTV" ){
      curr_dvh_name <- as.character((dvh))
    } else {
      curr_dvh_name <- as.character(str_to_title(dvh))
    }
      
    # selecting data of current DVH
    curr_dvh <- plans[[plan]][["DVH data"]][,idx]
      
    # adding current dvh to the dvh list
    dvh_list[[plan]][[curr_dvh_name]] <- curr_dvh
    
    dvh_list[[plan]] <- as.data.frame(dvh_list[[plan]])
    
    # gathering together dvh to plot 
    dvh_df[[plan]]  <- dvh_list[[plan]]  %>% 
      gather(key = "DHV_curve", value = "value", -Dose)
    
    # add label to the plan
    dvh_df[[plan]]$Configuration <- plan
    
  }
  
  # create plot grid
  plot <- ggplot(dvh_df[[plan_names[1]]], aes(x=Dose, y=value)) + 
    ylim(0,105) + 
    xlab("Dose [%]") +
    ylab("Volume [%]") 
  
  if(title == TRUE){
    # create plot title
    title <- c()
    for(i in 1:length(plan_names)){
      if(i==length(plan_names)){
        title <- paste(title, plan_names[i], sep=" ")
      } else{
        title <- paste(title, plan_names[i], "vs.", sep=" ")
      }
    }
    plot <- plot + ggtitle(title)
  } else if(title == FALSE){
    plot <- plot
  } else{
    plot <- plot + ggtitle(title)
  }
  
  # plot dvh for each plan
  for(plan in plan_names){
    if(str_detect(plan, "underdose")==TRUE || str_detect(plan, "overdose")==TRUE){
      plot <- plot + geom_line(data=dvh_df[[plan]], aes(color = Configuration), linewidth=0.4)
    }else{
      plot <- plot + geom_line(data=dvh_df[[plan]], aes(color = Configuration), linewidth=0.8)
    }
  }
  
  return(plot)
  
}

# ----------------------------------------------------------------------------- #

# upload comparison table output from FIonA

get_comparison <- function(table_csv, p1_name, p2_name){
  
  # ---------------------------------------------------- #
  # table_csv <- csv file of comparison, FIonA output
  # p1_name [chr] <- name of plan 1
  # p2_name [chr] <- name of plan 2
  # ---------------------------------------------------- #
  
  comparison <- read.csv(table_csv)
  
  colnames(comparison) <- c("Structure", paste("Volume [cc] - ", p1_name), paste("Volume [cc] - ", p2_name), paste("DMin [%] - ", p1_name), paste("DMin [%] - ", p2_name), paste("DMean [%] - ", p1_name), paste("DMean [%] - ", p2_name), paste("DMax [%] - ", p1_name), paste("DMax [%] - ", p2_name), paste("V95 [%] - ", p1_name), paste("V95 [%] - ", p2_name), paste("D98 [%] - ", p1_name), paste("D98 [%] - ", p2_name), paste("D5-D95 [%] - ", p1_name), paste("D5-D95 [%] - ", p2_name))
  
  return(comparison)

  }

# ----------------------------------------------------------------------------- #

# Read robustness table output from FIonA

read_robustness <- function(csv_file, organs){
  
  # ---------------------------------------------------- #
  # csv_file <- csv file of robustness, FIonA output
  # organs [chr] <- vector of organs of your plan
  # ---------------------------------------------------- #
  
  # read csv file
  robustness <- read.csv(csv_file)
  
  # get DVH names
  dvhs <- c("PTV", "CTV", toupper(organs))
  
  # rename first column as "Dose
  colnames(robustness)[1] <- "Dose"
  
  # rename columns of robustness table assigning correct numbers (num 5 is nominal)
  for(i in 1:length(dvhs)){
    
    # get column idxs for current DVH
    curr_idxs <- which(str_detect(colnames(robustness), dvhs[i]))
    
    # rename each column assigning correct numbers
    for(j in 1:length(curr_idxs)){
      colnames(robustness)[curr_idxs[j]] <- paste(dvhs[i], j, sep="_")
    }
    
  }
  
  # find NA values and substitute with 0
  robustness[is.na(robustness)] <- 0   
  rm(i,j,dvhs)
  
  return(robustness)
}

# ----------------------------------------------------------------------------- #

find_robustness_spread <- function(r){
  
  # ----------------------------------------------------------------------------------- #
  # r [table/dataframe] <- robustness table/df for which you want to find the robustness spread
  # ----------------------------------------------------------------------------------- #
  
  # get Dose values from the robustness table
  r_spread <- r$Dose
  r_spread <- as.data.frame(r_spread)
  colnames(r_spread) <- "Dose"
  
  # get DVH names 
  dvhs <- colnames(r)
  for(i in 1:length(dvhs)){
    dvhs[i] <- str_split(dvhs[i], pattern="_")[[1]][1]
  }
  dvhs <- unique(dvhs)
  dvhs <- dvhs[-1]
  
  # loop over each DVH to find minima and maxima of each DVH
  for(dvh in dvhs){
    
    # get dataframe of current DVH from the robustness table
    curr_dvh_idxs <- which(str_detect(colnames(r), dvh))
    curr_dvh <- r[curr_dvh_idxs[1]:tail(curr_dvh_idxs,n=1)]
    
    curr_dvh_max <- c()
    curr_dvh_min <- c()
    
    # get maxima for current DVH
    for(idx in 1:length(r$Dose)){
      curr_dvh_max[idx] <- max(curr_dvh[idx,])
    }
    
    # get minima for current DVH
    for(idx in 1:length(r$Dose)){
      curr_dvh_min[idx] <- min(curr_dvh[idx,])
    }
    
    # add columns of nominal, minima and maxima values of current DVH to spread table
    r_spread[[paste(dvh, "_nom", sep="")]] <- curr_dvh[[paste(dvh,"_5",sep="")]]
    r_spread[[paste(dvh, "_max", sep="")]] <- curr_dvh_max
    r_spread[[paste(dvh, "_min", sep="")]] <- curr_dvh_min

    rm(curr_dvh_idxs,curr_dvh,curr_dvh_max,curr_dvh_min)
  }
  
  return(r_spread)
  
}

# ----------------------------------------------------------------------------- #


plot_robustness <- function(robustness, plan_names, dvhs_stat, title){
  
  # ----------------------------------------------------------------------------------- #
  # robustness [list] <- list of robustness tables/dfs for which you want to plot the robustness spread
  # plan_names [chr] <- vector with names of plans for which you want to plot the robustness
  # dvhs_stat [logic] <- vector with logical conditions for DVHs to plot (either TRUE or FALSE)
  # ----------------------------------------------------------------------------------- #
  
  # for each robustness table in the input list, find robustness spread
  for(r in 1:length(robustness)){
    robustness[[r]] <- find_robustness_spread(robustness[[r]])
  }
  
  # set plan names
  names(robustness) <- plan_names
  dvhs <- toupper(names(dvhs_stat))
  
  dvhs_nom <- list()
  dvhs_min <- list()
  dvhs_max <- list()
  
  nom_df <- list()
  min_df <- list()
  max_df <- list()
  
  # loop over each plan to select just wanted DVHs
  for(plan in plan_names){
    
    # create nominal DVH list
    dvhs_nom[[plan]] <- list()
    dvhs_nom[[plan]][["Dose"]] <- robustness[[plan]]$Dose
    
    # create minima DVH list
    dvhs_min[[plan]] <- list()
    dvhs_min[[plan]][["Dose"]] <- robustness[[plan]]$Dose
    
    # create maxima DVH list
    dvhs_max[[plan]] <- list()
    dvhs_max[[plan]][["Dose"]] <- robustness[[plan]]$Dose
    
    # loop over DVHs to select the TRUE ones
    for(dvh_idx in 1:length(dvhs)){
      
      # if the organ is TRUE, I select the correspondent values
      if(dvhs_stat[[dvh_idx]]==TRUE){
        
        # select position of the organ in the dvh dataframe
        idx_nom <- which(str_detect(colnames(robustness[[plan]]), paste(dvhs[dvh_idx],"_nom",sep="")))
        idx_min <- which(str_detect(colnames(robustness[[plan]]), paste(dvhs[dvh_idx],"_min",sep="")))
        idx_max <- which(str_detect(colnames(robustness[[plan]]), paste(dvhs[dvh_idx],"_max",sep="")))
        
        # selecting current dvh in the dataframe and current name
        if(dvhs[dvh_idx] == "PTV" || dvhs[dvh_idx] == "CTV" ){
          curr_dvh_name <- as.character((dvhs[dvh_idx]))
        } else {
          curr_dvh_name <- as.character(str_to_title(dvhs[dvh_idx]))
        }
        
        # adding current nominal, minima and maxima DVHs to previous lists
        dvhs_nom[[plan]][[curr_dvh_name]] <- robustness[[plan]][,idx_nom]
        dvhs_min[[plan]][[curr_dvh_name]] <- robustness[[plan]][,idx_min]
        dvhs_max[[plan]][[curr_dvh_name]] <- robustness[[plan]][,idx_max]
      }
    }
    
    # transforming lists into dataframes
    dvhs_nom[[plan]] <- as.data.frame(dvhs_nom[[plan]])
    dvhs_min[[plan]] <- as.data.frame(dvhs_min[[plan]])
    dvhs_max[[plan]] <- as.data.frame(dvhs_max[[plan]])
    
    # gather nominal, minima and maxima curves together
    nom_df[[plan]]  <- dvhs_nom[[plan]]  %>% 
      gather(key = "DVH_nom", value = "value", -Dose)
    min_df[[plan]]  <- dvhs_min[[plan]]  %>% 
      gather(key = "DVH_min", value = "value", -Dose)
    max_df[[plan]]  <- dvhs_max[[plan]]  %>% 
      gather(key = "DVH_max", value = "value", -Dose)
    
    # adding plan label
    nom_df[[plan]]$Plan <- plan
    min_df[[plan]]$Plan <- plan
    max_df[[plan]]$Plan <- plan
    
    # adding type label
    nom_df[[plan]]$Curve <- "Nominal"
    min_df[[plan]]$Curve <- "Spread"
    max_df[[plan]]$Curve <- "Spread"
    
    # adding plan and type label
    nom_df[[plan]]$Plan_Curve <- paste(nom_df[[plan]]$Plan, nom_df[[plan]]$DVH_nom, sep=" - ")
    min_df[[plan]]$Plan_Curve <- paste(min_df[[plan]]$Plan, min_df[[plan]]$DVH_min, sep=" - ")
    max_df[[plan]]$Plan_Curve <- paste(max_df[[plan]]$Plan, max_df[[plan]]$DVH_max, sep=" - ")
    
  }
  
  colors <- c("cornflowerblue","violetred", "orange", "chartreuse","deeppink3", "darkturquoise", "mediumorchid1", "gold")
  
  # creating basic grid for plot
  plot <- ggplot(nom_df[[plan_names[1]]], aes(x=Dose, y=value)) + 
    ylim(0,105) + 
    xlab("Dose [%]") +
    ylab("Volume [%]")
  
  # creating title for the plot
  if(title == TRUE){
    # create plot title
    title <- c()
    for(i in 1:length(plan_names)){
      if(i==length(plan_names)){
        title <- paste(title, plan_names[i], sep=" ")
      } else{
        title <- paste(title, plan_names[i], "vs.", sep=" ")
      }
    }
    plot <- plot + ggtitle(title)
  } else if(title == FALSE){
    plot <- plot
  } else{
    plot <- plot + ggtitle(title)
  }
  
  # plotting DVHs for each plan
  for(plan in plan_names){
    plot <- plot + geom_line(data=nom_df[[plan]], aes(color = Plan_Curve, linetype=Curve), linewidth=0.8) +
      geom_line(data=min_df[[plan]], aes(color = Plan_Curve, linetype=Curve), linewidth=0.4) +
      geom_line(data=max_df[[plan]], aes(color = Plan_Curve, linetype=Curve), linewidth=0.4)
  }
  
  #plot <- plot + scale_color_manual(name = "DVH curves", values = colors)
  
  return(plot)
  
}


# ---------------------------------------------------------------------------- #

# get energies for each field and total number of spot for the current plan

get_energies <- function(fields.csv, fields_num, tot_spots){
  
  # -------------------------------------------------------------------------- #
  # fields.csv [chr] <- vector with csv file names for each field
  # fields_num [chr] <- vector with field numbers
  # -------------------------------------------------------------------------- #
  
  energies <- list()

  for(field in 1:length(fields.csv)){
    
    curr_field_df <- read.csv(fields.csv[field], sep=" ", header = FALSE)
    colnames(curr_field_df) <- c("n_spots", "energy")
    curr_field <- list()
    curr_field[["Energy values"]] <- unique(curr_field_df$energy)
    curr_field[["Number of energies"]] <- length(unique(curr_field_df$energy))
    energies[[fields_num[field]]] <- curr_field
    rm(curr_field_df)
    
  }
  
  energies[["Tot. spots"]] <- tot_spots
  
  return(energies)
  
}

# ----------------------------------------------------------------------------- #

# calculates the dead time to change spot position and energy value for a plan

dead_time <- function(TS, TE, plan_energies){
  
  # ---------------------------------------------------------- #
  # TS [num] <- time [ms] to change position between two spots
  # TE [num] <- time [ms] to change energy between two layers
  # plan_energies <- output of get_energies for a plan
  # ---------------------------------------------------------- #
  
  # calculate dead time for spot positioning
  ts <- TS*plan_energies[["Tot. spots"]]
  
  # calculate dead time for energy change
  fields <- names(plan_energies)
  rm_last <- which(fields == tail(fields,n=1))
  fields <- fields[-rm_last]
  
  te <- 0
  for(field in fields){
    te <- te + TE*p1_energies[[field]][["Number of energies"]]
  }
  
  # converting dead time from ms to s
  dt <- (ts+te)*1e-3
  
  return(dt)
  
}



```



In this pipeline we are going to compare different plans for proton therapy, either with or without the use of a Ridge Filter.
```{r}
setwd("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5")

plan1 <- read_plan("Plan 1/dvhs_nominal.csv")
plan2 <- read_plan("Plan 2/dvhs_nominal.csv")
plan3 <- read_plan("Plan 3/dvhs_nominal.csv")
```

The organs at risk (OAR) involved in this case are:
```{r}
cat("\n\n")
get_organs(plan = plan1)
```


# Plan 1

The first plan is composed by two fields F0 and F1 was calculated by using the following settings:

- target: PTV
- preabsorber: Auto
- nozzle: 15

for both fields F0 and F1, and:

- gantry angle for F0: 130
- couch angle for F0: 180
- gantry angle for F1: 110
- couch angle for F1: 0


```{r}
plot_DVHs(plans = list(plan1), plan_names = "Plan 1", dvhs_stat = c(PTV=TRUE, CTV=TRUE, LIVER=TRUE, STOMACH=TRUE), title = "Plan 1 - no RF") 
ggsave()

```

# Plan 2

The second plan is also composed by two fields F0 and F1 and was calculated by using the following settings:

- target: PTV
- preabsorber: Auto
- nozzle: 15

for both fields F0 and F1, and:

- gantry angle for F0: 45
- couch angle for F0: 180
- gantry angle for F1: 55
- couch angle for F1: 0

```{r}
plot_DVHs(plans = list(plan2), plan_names = "Plan 2", dvhs_stat = c(PTV=TRUE, CTV=TRUE, LIVER=TRUE, STOMACH=TRUE), title = TRUE) 
```

# Plan 1 and Plan 2 comparison

We can compare the two plans on the same plot, either by plotting all the DVHs or by selecting some of them:

```{r}
plot_DVHs(plans = list(plan1,plan2), plan_names = c("Plan 1", "Plan 2"), dvhs_stat = c(PTV=TRUE, CTV=TRUE, LIVER=TRUE, STOMACH=TRUE), title = TRUE) 

plot_DVHs(plans = list(plan1,plan2), plan_names = c("Plan 1", "Plan 2"), dvhs_stat = c(PTV=TRUE, CTV=FALSE, LIVER=TRUE, STOMACH=FALSE), title = TRUE) 
```

As we can see from the DVHs in the plots above, Plan 1 is better than Plan 2 since the DVHs for the OAR are much lower, meaning that we have less dose delivered to these organs.

We can visually see that by looking at the following plans' screenshots:

## Visual comparison between Plan 1 and Plan 2 - heat map
```{r}
setwd("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5")

img <- readPNG(source="Plans comparison/P1_P2_comparison.png")
grid::grid.raster(img)
```

## Visual comparison between Plan 1 and Plan 2 - target coverage map
```{r}
setwd("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5")

img <- readPNG(source="Plans comparison/P1_P2_comparison_target_coverage.png")
grid::grid.raster(img)
```


## DVHs characteristics
To better analyze the differences between these two plans, we can compare them by looking at the following table:

```{r}
setwd("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5")

df <- get_comparison("Plans comparison/statistics_P1_p2.csv", "Plan 1", "Plan 2")
kable(df)
```

By looking at the values of V95 for CTV and D98 for PTV, we could say at first that Plan 2 is better than Plan 1, since both these values for Plan 2 are higher than for Plan 1. 

Indeed, V95 = 100.00 indicates that the whole CTV volume in Plan 2 receives at least 95% of the dose, meanwhile for Plan 1 we have V95 = 95.27, indicating that only 95.27% of the CTV volume receives 95% of the dose. 
The same considerations can be made also for D98, which indicates the minimum dose that 98% of the volume of an organ receives. 
These are both measures of the target coverage.

However, if we look at the mean dose received by the organs we can clearly say that Plan 1 is better than Plan 2, since in the latter the mean dose delivered to the OAR is higher than in the former.
Moreover, if we look at the values of D5-D95 of the PTV, we could say that Plan 1 has a higher homogeneity in the delivered dose, since this value is smaller than the one for Plan 2, indicating a steeper dose fall-off.


## Robustness analysis
We can next visualize the output from robustness analysis.

Also in this case we can select the DVHs of interest and compare more than one plan.

```{r}
setwd("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5")

cat("\n\n")
r1 <- read_robustness("Plan 1/dvhs_robustness.csv", organs = c("liver","stomach"))
r2 <- read_robustness("Plan 2/dvhs_robustness.csv", organs = c("liver","stomach"))
r <- list(r1,r2)
names(r) <- c("Plan 1", "Plan 2")
```


```{r}
plot_robustness(robustness = list(r1), plan_names = c("Plan 1"), dvhs_stat = c(PTV=TRUE,CTV=TRUE,STOMACH=TRUE,LIVER=TRUE), title = FALSE)
```


```{r}
plot_robustness(robustness = list(r2), plan_names = c("Plan 2"), dvhs_stat = c(PTV=TRUE,CTV=TRUE,STOMACH=TRUE,LIVER=TRUE), title = FALSE)
```

From these plots, we could say that Plan 2 seems more robust than Plan 1, at least for what it concerns the OAR, since the spread around the nominal curves is smaller. This is very clear for example for the liver, as shown below:

```{r}
plot_robustness(robustness = list(r1,r2), plan_names = c("Plan 1", "Plan 2"), dvhs_stat = c(PTV=FALSE,CTV=FALSE,STOMACH=FALSE,LIVER=TRUE), title = FALSE)
```

Finally, we could also recalculate the dose to see the underdose and overdose for the two plans.

```{r}
setwd("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5")

p1_overdose <- read_plan("Plan 1/dvhs_overdose.csv")
p1_underdose <- read_plan("Plan 1/dvhs_underdose.csv")

p2_overdose <- read_plan("Plan 2/dvhs_overdose.csv")
p2_underdose <- read_plan("Plan 2/dvhs_underdose.csv")

plot_DVHs(plans = list(plan1,p1_overdose,p1_underdose), plan_names = c("Plan 1 - nominal", "Plan 1 - overdose", "Plan 1 - underdose"), dvhs_stat = c(PTV=TRUE,CTV=TRUE,STOMACH=TRUE,LIVER=TRUE), title = "Plan 1 - Underdose vs. Overdose")

plot_DVHs(plans = list(plan2,p2_overdose,p2_underdose), plan_names = c("Plan 2 - nominal", "Plan 2 - overdose", "Plan 2 - underdose"), dvhs_stat = c(PTV=TRUE,CTV=TRUE,STOMACH=TRUE,LIVER=TRUE), title = "Plan 2 - Underdose vs. Overdose")

```

In this case, instead, we can visually see that Plan 1 gives us better results for the OAR since the deviation of the overdose and the underdose form the nominal dose distribution is not as large as for Plan 2, meanwhile for the CTV Plan 2 is better.

# Plan 3 - Ridge Filter

We now analyze a plan with the use of a Ridge Filter (RF).

This plan is composed by two fields F0 and F1 which were calculated by using the same setting as Plan 1, with the only difference that in this case $\Delta U = \Delta T = 0.47$, since, as the beam passes through the filter, it undergoes scattering and increases in size.

The DVHs of this plan are shown below:
```{r}
plot_DVHs(plans = list(plan3), plan_names = "Plan 3", dvhs_stat = c(PTV=TRUE, CTV=TRUE, LIVER=TRUE, STOMACH=TRUE), title = TRUE) 
```

We would like to compare Plan 1 with Plan 3 and see which are the differences in using a Ridge Filter with pencil beam scanning (PBS). In particular, we would like to count the number of spots and the number of energies in the two cases.

The number of spots and energy layers used in Plan 1 is:
```{r}
setwd("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5")

p1_fields.csv <- c("Plan 1/energies_F0.csv","Plan 1/energies_F1.csv")
p1_fields_num <- c("F0","F1")
p1_energies <- get_energies(p1_fields.csv, p1_fields_num, tot_spots = 14435)

print(paste("Number of spots:",p1_energies[["Tot. spots"]]))
print(paste("Number of energy layers in F0:",p1_energies[["F0"]][["Number of energies"]]))
print(paste("Number of energy layers in F1:",p1_energies[["F1"]][["Number of energies"]]))

```

meanwhile, the number of spots used in Plan 3 is:
```{r}
setwd("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5")

p3_fields.csv <- c("Plan 3/energies_F0.csv","Plan 3/energies_F1.csv")
p3_fields_num <- c("F0","F1")
p3_energies <- get_energies(p3_fields.csv, p3_fields_num, tot_spots = 1553)

print(paste("Number of spots:",p3_energies[["Tot. spots"]]))
print(paste("Number of energy layers in F0:",p3_energies[["F0"]][["Number of energies"]]))
print(paste("Number of energy layers in F1:",p3_energies[["F1"]][["Number of energies"]]))
```

As we can see, the use of the RF leads to two main benefits:

- the decrease of the number of energy layers, due to the fact that we create a small spread-out Bragg Peak (SOBP), thus spreading the energy range of the beam arriving at a certain spot position 

- the decrease of the number of spots, due to the fact that the beam enlarges a bit when traversing the filter and due to the reduction of the number of energy layers

In this way, by using a RF, we can reduce the plan delivery time to the patient. Indeed, if we measure the dead time of the two plans and we compare them, if we consider a spot positioning time of about 2ms and an energy adjustment time of about 150ms, we get:
```{r}
TS = 2
TE = 150
dt_1 <- dead_time(TS,TE,p1_energies)
dt_3 <- dead_time(TS,TE,p3_energies)

print(paste("Dad time for Plan 1:", dt_1, "s"))
print(paste("Dad time for Plan 3:", dt_3, "s"))
```

which corresponds to a time reduction percentage of about:
```{r}
(dt_1-dt_3)/dt_1
```

Below, we show the difference between the DVHs for Plan 1 and Plan 3:
```{r}
plot_DVHs(plans = list(plan1,plan3), plan_names = c("Plan 1", "Plan 3"), dvhs_stat = c(PTV=TRUE, CTV=TRUE, LIVER=TRUE, STOMACH=TRUE), title = TRUE) 
```

and we plot the results of the robustness analysis for both plans:
```{r}
setwd("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5")

cat("\n\n")
r1 <- read_robustness("Plan 1/dvhs_robustness.csv", organs = c("liver","stomach"))
r3 <- read_robustness("Plan 3/dvhs_robustness.csv", organs = c("liver","stomach"))
r <- list(r1,r3)
names(r) <- c("Plan 1", "Plan 3")

plot_robustness(robustness = list(r1), plan_names = c("Plan 1"), dvhs_stat = c(PTV=TRUE,CTV=TRUE,STOMACH=TRUE,LIVER=TRUE), title = TRUE)

plot_robustness(robustness = list(r3), plan_names = c("Plan 3"), dvhs_stat = c(PTV=TRUE,CTV=TRUE,STOMACH=TRUE,LIVER=TRUE), title = TRUE)
```

From these plots, we can see that the robustness of the plans is quite similar. The main difference that can be seen is that the planned dose to be delivered to the liver is more robust for Plan 3 than for Plan 1, as we can better see below:
```{r}
plot_robustness(robustness = list(r1,r3), plan_names = c("Plan 1", "Plan 3"), dvhs_stat = c(PTV=FALSE,CTV=FALSE,STOMACH=FALSE,LIVER=TRUE), title = TRUE)
```

???

```{r}
plot_robustness(robustness = list(r1,r3), plan_names = c("Plan 1", "Plan 3"), dvhs_stat = c(PTV=FALSE,CTV=FALSE,STOMACH=TRUE,LIVER=FALSE), title = TRUE)
```

# Set up different configurations with the Ridge Filter
We tried different configurations of spot placement for our set up with the Ridge Filter.

We arranged spots by considering:

- spot spacing based on spot size in air, water or on a fixed grid with $\Delta U = \Delta T = 0.47 cm$
- spot positioning using an hexagonal or a rectilinear grid
- constant layer spacing factor of 3

We have thus six different configurations:
- Rectilinear fixed grid with $\Delta U = \Delta T = 0.47 cm$
- Hexagonal fixed grid with with $\Delta U = \Delta T = 0.47 cm$
- Rectilinear grid with spot spacing in air with a spot spacing factor of 1.2
- Hexagonal grid with spot spacing in air with a spot spacing factor of 1.2
- Rectilinear grid with spot spacing in water with a spot spacing factor of 1.5
- Hexagonal grid with spot spacing in water with a spot spacing factor of 1.5

The difference in the spot spacing factor for the spot placings in water and in air is due to che choice we made for the best optimized plan that gave a us a better result of PTV V95.

Below is shown the table with the configuration parameters and the values of PTV V95 and the number of spots obtained for each of the six configurations.
```{r}
# read optimization table

# opt_table <- read.csv("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5/best_opt.csv", header=TRUE, sep=";")

opt_table <- read_excel("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5/best_opt.xlsx")
opt_table <- as.data.frame(opt_table)

#names(opt_table[,2]) <- "Number of spots"
#names(opt_table[,3]) <- "PTV V95 [%]"
#names(opt_table[,4]) <- "Sigma [cm]"
#names(opt_table[,5]) <- "Margin [cm]"
#names(opt_table[,6]) <- "Dose at margin [%]"
#names(opt_table[,8]) <- "Structure margin [cm]"

kable(opt_table)
```
Below is shown the values of PTV V95 vs. the number of spots for each configuration.
```{r}

df_v95_nspots <- data_frame("PTV_V95" = as.numeric(opt_table[,3]), "n_spots" = as.numeric(opt_table[,2]), "Configuration" = opt_table[,1])

# plot the PTV V95 vs. spots
ggplot(df_v95_nspots, aes(x=PTV_V95, y=n_spots)) + 
  geom_point(aes(x=PTV_V95, y=n_spots, color=Configuration), size=4) +
  xlab("PTV V95 [%]") +
  ylab("Number of Spots")
```

As we can see from this plot, the best configurations which give us the highest values of PTV V95 and the lowest number of spots is the one with spot placing using spot size in water.

To further analyze these results, we can plot the DVHs for the PTV and the organs.

```{r}
# load dvhs for each configuration
R_grid <- read_plan("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5/DVHs optimization/Rgrid.csv")
H_grid <- read_plan("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5/DVHs optimization/Hgrid.csv")
R_air <- read_plan("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5/DVHs optimization/Rair.csv")
H_air <- read_plan("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5/DVHs optimization/Hair.csv")
R_water <- read_plan("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5/DVHs optimization/Rwat.csv")
H_water <- read_plan("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5/DVHs optimization/Hwat.csv")

plans <- list(R_grid, H_grid, R_air, H_air, R_water, H_water)
plan_names <- c("R grid", "H grid", "R air (ssf = 1.2)", "H air (ssf = 1.2)", "R water (ssf = 1.5)", "H water (ssf = 1.5)")

PTV <- plot_DVH_comparison(plans = plans, plan_names = plan_names, dvh = "PTV", title = "PTV comparison for the different configurations")
PTV + coord_cartesian(xlim=c(85,120))
```

As we can see from this plot, the PTV DVHs of the two configurations using spot placement in water are overlapped and are the steepest ones compared to others.

If we look instead to the DVHs for the two organs, we can see that the best curves are actually not from the spot placement in water configurations. These are actually the "worst":

```{r}
liver <- plot_DVH_comparison(plans = plans, plan_names = plan_names, dvh = "liver", title = "Liver DVHs comparison for the different configurations")
liver + coord_cartesian(ylim=c(0,20))

stomach <- plot_DVH_comparison(plans = plans, plan_names = plan_names, dvh = "stomach", title = "Stomach DVHs comparison for the different configurations")
stomach + coord_cartesian(ylim=c(0,25))
```

Finally, we can visually see the robustness of the CTV for each configuration.

```{r}
# robustness spread
CTV_Rgrid <- read_robustness("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5/DVHs robustness/CTV_Rgrid.csv", organs = c())
CTV_Hgrid <- read_robustness("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5/DVHs robustness/CTV_Hgrid.csv", organs = c())
CTV_Rair <- read_robustness("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5/DVHs robustness/CTV_Rair.csv", organs = c())
CTV_Hair <- read_robustness("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5/DVHs robustness/CTV_Hair.csv", organs = c())
CTV_Rwat <- read_robustness("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5/DVHs robustness/CTV_Rwat.csv", organs = c())
CTV_Hwat <- read_robustness("/Users/martinabonomi/Library/Mobile Documents/com~apple~CloudDocs/Planning/Liver5/DVHs robustness/CTV_Hwat.csv", organs = c())

CTV_rob <- plot_robustness(robustness = list(CTV_Rgrid, CTV_Hgrid, CTV_Rair, CTV_Hair, CTV_Rwat, CTV_Hwat), plan_names = c("R grid", "H grid", "R air", "H air", "R water", "H water"), dvhs_stat = c(PTV=FALSE, CTV=TRUE, LIVER=FALSE, STOMACH=FALSE), title = "CTV robustness spread")
CTV_rob + coord_cartesian(xlim=c(85,120))
```

As we can see, the plans are pretty robust since the spread around the nominal curve is very small. In this case, the steepest and best CTV is R water.
